maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-08-21 docs/paper/figures/tex/fig1_diagram.tex
fix minor typos

\begin{figure} 
	\begin{minipage}{5in}
	  \input{diagram/diagram.tikz} 
	\end{minipage}
	\caption{
	  Usage of maze-dataset. We create a \texttt{MazeDataset} from a \texttt{MazeDatasetConfig}. This contains \texttt{SolvedMaze} objects which can be converted to and from a variety of formats. Code in the image contains clickable links to \docslink{maze_dataset.html}{documentation}. A variety of generated examples can be viewed \docslink{examples/maze_examples.html}{here}.
	}
	\label{fig:diagram}
  \end{figure}

>>>>

\begin{figure} 
	\resizebox{\linewidth}{!}{
    	\input{diagram/diagram.tikz}
	}
	\caption{
	  Usage of maze-dataset. We create a \texttt{MazeDataset} from a \texttt{MazeDatasetConfig}. This contains \texttt{SolvedMaze} objects which can be converted to and from a variety of formats. A variety of generated examples can be viewed on the \docslink{examples/maze_examples.html}{examples page}, and more information can be found in the \docslink{maze_dataset.html}{documentation}.
	}
	\label{fig:diagram}
\end{figure}

### 2025-08-21 docs/paper/figures/tex/fig6_benchmarks.tex
fix minor typos

\begin{figure}
	\hypertarget{fig:benchmarks}{%
		\centering
		\includegraphics[width=0.9\textwidth]{figures/benchmarks/gridsize-vs-gentime.pdf}
		\caption{
			Plot of maze generation time. Generation time scales
			exponentially with maze size for all algorithms. Generation time per
			maze does not depend on the number of mazes being generated, and there
			is minimal overhead to initializing the generation process for a small
			dataset. Wilson's algorithm is notably less efficient than others and
			has high variance. Note that values are averaged across all parameter
			sets for that algorithm. More information can be found on the
			\href{https://understanding-search.github.io/maze-dataset/benchmarks/}{benchmarks page}.
		}
		\label{fig:benchmarks}
	}
\end{figure}

>>>>

\begin{figure}
	\hypertarget{fig:benchmarks}{%
		\centering
		\includegraphics[width=\textwidth]{figures/benchmarks/gridsize-vs-gentime.pdf}
		\caption{
			Plot of maze generation time. Generation time scales
			roughly exponentially with maze size for all algorithms. Generation time per
			maze does not depend on the number of mazes being generated, and there
			is minimal overhead to initializing the generation process for a small
			dataset. Wilson's algorithm is notably less efficient than others and
			has high variance. Note that values are averaged across all parameter
			sets for that algorithm. More information can be found on the
			\href{https://understanding-search.github.io/maze-dataset/benchmarks/}{benchmarks page}.
		}
		\label{fig:benchmarks}
	}
\end{figure}

### 2025-08-21 docs/paper/paper.tex
fix minor typos

While maze generation itself is straightforward, the architectural
challenge comes from building a system supporting many algorithms with
configurable parameters, property filtering, and representation
transformation. This library aims to greatly streamline the process of
generating and working with datasets of mazes that can be described as
subgraphs of an \(n \times n\) lattice with boolean connections and,
optionally, start and end points that are nodes in the graph.
Furthermore, we place emphasis on a wide variety of possible text output
formats aimed at evaluating the spatial reasoning capabilities of Large
Language Models (LLMs) and other text-based transformer models.

@@

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information and a tensor optimized for storing
sub-graphs of a lattice. These objects can be converted to and from
several formats, shown in \autoref{fig:output-fmts}, to maximize their
utility in different contexts.

@@

This package was originally built for the needs of the
(\citeproc{ref-maze-transformer-github}{Michael I. Ivanitskiy et al.,
2023b}) project, which aims to investigate spatial planning and world
models in autoregressive transformer models trained on mazes
(\citeproc{ref-ivanitskiy2023structuredworldreps}{Michael Igorevich
Ivanitskiy, Spies, et al., 2023};
\citeproc{ref-maze-dataset-arxiv-2023}{Michael Igorevich Ivanitskiy,
Shah, et al., 2023}; \citeproc{ref-spies2024causalworldmodels}{Spies et
al., 2024}). It was extended for work on understanding the mechanisms by
which recurrent convolutional and implicit networks
(\citeproc{ref-fung2022jfb}{Fung et al., 2022}) solve mazes given a
rasterized view (\citeproc{ref-knutson2024logicalextrapolation}{Knutson
et al., 2024}), which required matching the pixel-padded and endpoint
constrained output format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}). Ongoing work using
\texttt{maze-dataset} aims to investigate the effects of varying the
tokenization format on the performance of pretrained LLMs on spatial
reasoning.

This package has also been utilized in work by other groups:

\begin{itemize}
\item
  By (\citeproc{ref-nolte2024multistep}{Nolte et al., 2024}) to compare
  the effectiveness of transformers trained with the MLM-\(\mathcal{U}\)
  (\citeproc{ref-MLMU-kitouni2024factorization}{Kitouni et al., 2024})
  multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\citeproc{ref-wang2024imperative}{Wang et al., 2024}) and
  (\citeproc{ref-chen2024iaimperative}{Chen et al., 2024}) to study the
  effectiveness of imperative learning.
\item
  By (\citeproc{ref-zhang2025tscend}{Zhang et al., 2025}) to introduce a
  novel framework for reasoning diffusion models.
\item
  By (\citeproc{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to improve
  spatial reasoning in LLMs with GRPO.
\end{itemize}

@@

@@

@@

\bibitem[\citeproctext]{ref-zhang2025tscend}
Zhang, T., Pan, J.-S., Feng, R., \& Wu, T. (2025). T-SCEND: Test-time
scalable MCTS-enhanced diffusion model. \emph{arXiv Preprint
arXiv:2502.01989}. \url{https://doi.org/10.48550/arXiv.2502.01989}

>>>>

While maze generation itself is straightforward, the architectural
challenge comes from building a system supporting many algorithms with
configurable parameters, property filtering, representation
transformation, and reproducibility. This library aims to greatly
streamline the process of generating and working with datasets of mazes
that can be described as subgraphs of an \(n \times n\) lattice with
boolean connections and, optionally, start and end points that are nodes
in the graph. Furthermore, we place emphasis on a wide variety of
possible text output formats aimed at evaluating the spatial reasoning
capabilities of Large Language Models (LLMs) and other text-based
transformer models.

@@

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information and an array optimized for storing
sub-graphs of a lattice. These objects can be converted to and from
several formats, shown in \autoref{fig:output-fmts}, to maximize their
utility in different contexts.

@@

This package was originally built for the needs of the
\texttt{maze-transformer} project
(\citeproc{ref-maze-transformer-github}{Michael I. Ivanitskiy et al.,
2023b}), which aims to investigate spatial planning and world models in
autoregressive transformer models trained on mazes
(\citeproc{ref-ivanitskiy2023structuredworldreps}{Michael Igorevich
Ivanitskiy, Spies, et al., 2023};
\citeproc{ref-maze-dataset-arxiv-2023}{Michael Igorevich Ivanitskiy,
Shah, et al., 2023}; \citeproc{ref-spies2024causalworldmodels}{Spies et
al., 2024}). It was extended for work on understanding the mechanisms by
which recurrent convolutional and implicit networks
(\citeproc{ref-fung2022jfb}{Fung et al., 2022}) solve mazes given a
rasterized view (\citeproc{ref-knutson2024logicalextrapolation}{Knutson
et al., 2024}), which required matching the pixel-padded and endpoint
constrained output format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}). Ongoing work using
\texttt{maze-dataset} aims to investigate the effects of varying the
tokenization format on the performance of pretrained LLMs on spatial
reasoning.

At the time of writing, this software package has been actively used in
work by other groups:

\begin{itemize}
\item
  By (\citeproc{ref-nolte2024multistep}{Nolte et al., 2024}) to compare
  the effectiveness of transformers trained with the MLM-\(\mathcal{U}\)
  (\citeproc{ref-MLMU-kitouni2024factorization}{Kitouni et al., 2024})
  multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\citeproc{ref-wang2024imperative}{Wang et al., 2024}) and
  (\citeproc{ref-chen2024iaimperative}{Chen et al., 2024}) to study
  imperative learning.
\item
  By (\citeproc{ref-zhang2025tscend}{Zhang et al., 2025a}) to introduce
  a novel framework for reasoning diffusion models.
\item
  By (\citeproc{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to improve
  spatial reasoning in LLMs with GRPO.
\item
  By (\citeproc{ref-cai2025morse}{Cai et al., 2025}) to create a
  multimodal reasoning benchmark, via mazes in videos.
\item
  By (\citeproc{ref-xu2025visual}{Xu et al., 2025}) to study visual
  planning in LLMs.
\item
  By (\citeproc{ref-lee2025adaptive}{Lee et al., 2025}) to evaluate
  adaptive inference-time scaling with diffusion models on maze
  navigation tasks.
\item
  By (\citeproc{ref-zhang2025vfscale}{Zhang et al., 2025b}) to test
  verifier-free diffusion models.
\end{itemize}

@@

\bibitem[\citeproctext]{ref-cai2025morse}
Cai, Z., Wang, A., Satheesh, A., Nakhawa, A., Jae, H., Powell, K., Liu,
M., Jay, N., Oh, S., Wang, X., \& others. (2025). MORSE-500: A
programmatically controllable video benchmark to stress-test multimodal
reasoning. \emph{arXiv Preprint arXiv:2506.05523}.
\url{https://doi.org/10.48550/arXiv.2506.05523}

@@

\bibitem[\citeproctext]{ref-lee2025adaptive}
Lee, G., Bao, T. N. N., Yoon, J., Lee, D., Kim, M., Bengio, Y., \& Ahn,
S. (2025). Adaptive cyclic diffusion for inference scaling. \emph{arXiv
Preprint arXiv:2505.14036}.
\url{https://doi.org/10.48550/arXiv.2505.14036}

@@

\bibitem[\citeproctext]{ref-xu2025visual}
Xu, Y., Li, C., Zhou, H., Wan, X., Zhang, C., Korhonen, A., \& Vulić, I.
(2025). Visual planning: Let's think only with images. \emph{arXiv
Preprint arXiv:2505.11409}.
\url{https://doi.org/10.48550/arXiv.2505.11409}

\bibitem[\citeproctext]{ref-zhang2025tscend}
Zhang, T., Pan, J.-S., Feng, R., \& Wu, T. (2025a). T-SCEND: Test-time
scalable MCTS-enhanced diffusion model. \emph{arXiv Preprint
arXiv:2502.01989}. \url{https://doi.org/10.48550/arXiv.2502.01989}

\bibitem[\citeproctext]{ref-zhang2025vfscale}
Zhang, T., Pan, J.-S., Feng, R., \& Wu, T. (2025b). VFScale: Intrinsic
reasoning through verifier-free test-time scalable diffusion model.
\emph{arXiv Preprint arXiv:2502.01989}.
\url{https://doi.org/10.48550/arXiv.2502.01989}
